---
title: "R Notebook"
output: html_notebook
---

##Question 1

```{r}
install.package(zoo)
```


A decision tree can be built to decide if new subjects are likely to have a BMI>25 
or  not,  based  on  the  two  discrete-valued  variables  “Play  Football?”  and  “Watch 
Football?”, together with the continuous variable “age”.  
a) Using the criterion of entropy-based information gain, determine the best split-
point of the continuous variable “age” at the root node of the decision tree. 

```{r echo=, warning=FALSE}
library(zoo)
playFootbal <- c(4, 5)
watchFootball <- c(5, 4)
age <- c(25, 27, 28, 23, 27, 24, 25, 21, 21)
age <- sort(age)
age <-  rollmean(age, 2)
Enthropy <- function(n, m , k = 2){
  
}
informationGain <- function(df, splitPoint, k = 2){
  
  ES <- -(4/9)*log(4/9, k)-(5/9)*log(5/9, 2)
  E
  result <- 
}


```

b) Hence  determine  which  of  the  three  variables:  “Play  Football?”,  “Watch 
Football?”, and “age”, is the most appropriate to place as the decision variable 
at the root node of the decision tree.  

```{r echo=FALSE, warning = F}

```

